---
title: "Reproducible Research Course Project 1"
author: "David R"
date: "30/04/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the first course project for the Reproducible Research course.  
First, the following [repository](https://github.com/rdpeng/RepData_PeerAssessment1) was cloned containing the original dataset *activity.zip*. Then the csv file was manually unzipped and stored as *activity.csv* in the same folder.  


## Loading the data

Load the data via the *read.csv* function. 

```{r}
activitydata<-read.csv("activity.csv")
```

## Calculate mean number of steps taken per day

First load dplyr package to make use of pipeline operator and functions such as 
group_by, summarize. 

```{r}
library(dplyr)
```

Next, sum totalsteps by day and store in new variable *aggActivity*.  
Then calculate mean of total steps per day. 

```{r}

activitydata %>%as_tibble() %>% 
                group_by(date) %>% 
                summarize(totalsteps=sum(steps)) -> aggActivity

mean(aggActivity$totalsteps, na.rm=TRUE)
```
and same for the median: 
```{r}
median(aggActivity$totalsteps, na.rm=TRUE)
```


